---
title: "p8105_hw2_dt2833"
author: "Dominique Tzou"
date: "2026-10-04"
output: github_document


---


# Question 1


```{r}
library(tidyverse)
library(dplyr)

transit_df = read_csv(file = "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
 na = c("NA", ".", ""))|>
janitor::clean_names() |>
select (line, station_name, station_latitude, station_longitude, route1,route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, entry, vending, entrance_type, ada)|> 
  distinct() |>
  mutate(
    entry = recode_values(
      entry,
      "YES"  ~ TRUE,
      "NO" ~ FALSE
    )) #case_match()` was deprecated in dplyr 1.2.0.
```


```{r ada}
transit_distinct_df = 
  distinct(transit_df, station_name, line, .keep_all = TRUE) 

transit_distinct_df|> 
filter(ada ==TRUE) |> 
  nrow()
```


```{r vending and entry}
vending_no = filter(transit_df,vending == "NO")

filter(transit_df,vending == "NO")|> 
  nrow() #117

vending_no_nrow = vending_no |>nrow()

filter(transit_df,vending == "NO", entry == "TRUE") |>
  nrow() #45

vending_no_entry_yes = filter(vending_no, entry == "TRUE") |>
  nrow() #45
```


```{r reformat routes}
#how many a train stations
a_train = 
  transit_df |>
  mutate (
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)
    ) |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number", 
    names_prefix = "route",
    values_to = "route_name") |>
  
filter(route_name=="A" )|>
 distinct( station_name, line, .keep_all = TRUE) |>
  nrow() #60

#how many are ada compliant
ada_atrain = 
transit_df |>
  mutate (
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)
    ) |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number", 
    names_prefix = "route",
    values_to = "route_name") |>
  
filter(route_name=="A" )|>
 distinct( station_name, line, .keep_all = TRUE) |>
filter(ada=="TRUE" )|>
nrow()

```

__Description__

Dataset contains variables: `r names(transit_df)`. Each route variable represents an additional line that station serves (the most train routes served at one station is 11). For data cleaning, we have cleaned names, selected our desired variables, and standardized missing values to NA. This dataset contains `r ncol(transit_df)` columns and `r nrow(transit_df)` rows. 

How many distinct stations are there?  `r distinct(transit_df, station_name, line, .keep_all = TRUE)  |> nrow()` unique stations 

This dataset could be more "tidier", the 11 route variable columns are difficult to read . 


How many stations are ADA compliant? `r transit_distinct_df|> filter(ada ==TRUE) |> nrow() ` stations

What proportion of station entrances / exits without vending allow entrance?
`r filter(transit_df,vending == "NO", entry == "TRUE") |>nrow()`  / `r filter(transit_df,vending == "NO")|> nrow()` = `r ((filter(transit_df,vending == "NO", entry == "TRUE") |>nrow() / filter(transit_df,vending == "NO")|> nrow())*100) |> round(digits = 2)` % station entrances. 

How many distinct stations serve the A train? `r a_train` stations. 

Of the stations that serve the A train, how many are ADA compliant? `r ada_atrain` stations. 

#Question 2 

Importing data from Mr. Trash Wheel, Professor Trash Wheel, Gwynnda Trash Wheel.
```{r import data}
library(readxl)

mrtrash_df = read_excel("./data/202610 Trash Wheel Collection Data.xlsx", sheet="Mr. Trash Wheel",  
             na = c("NA", ".", ""))|>
             janitor::clean_names() 
  
proftrash_df = read_excel("./data/202610 Trash Wheel Collection Data.xlsx", sheet="Professor Trash Wheel",
              na = c("NA", ".", "", "dive"))|>
             janitor::clean_names() 

gwynndatrash_df = read_excel("./data/202610 Trash Wheel Collection Data.xlsx", sheet="Gwynnda the Good Wheel of the W",
              na = c("NA", ".", ""))|>
             janitor::clean_names()  
```

Cleaning Mr. Trash Wheel
```{r}
mrtrash_df = 
  select(mrtrash_df,-x15, -x16)|>  #deleted unnecessary columns 
  filter (!is.na(dumpster)) |> #deleted non-dumpster specific data rows
  mutate(
        sports_balls = round(sports_balls), 
        sports_balls = as.integer(sports_balls), 
        trash_wheel = "Mr. Trash Wheel") |> # creating new variable to indicate wheel source of data 
  relocate(trash_wheel) 

```

Cleaning Professor Trash Wheel
```{r Professor Trash Wheel}
proftrash_df =
  filter (proftrash_df, !is.na(dumpster)) |> #deleted non-dumpster specific data rows
  mutate(trash_wheel = "Professor Trash Wheel")  |> # creating new variable to indicate wheel source of data 
  relocate(trash_wheel) 
```

Cleaning Gwynnda Trash Wheel
```{r Gwynnda Trash Wheel}
gwynndatrash_df = 
  filter (gwynndatrash_df, !is.na(dumpster)) |> #deleted non-dumpster specific data rows
  mutate(trash_wheel = "Gwynnda Trash Wheel") |> # creating new variable to indicate wheel source of data 
  relocate(trash_wheel) 

```

Combining into 1 dataframe
```{r}

str(mrtrash_df) # year variable needs to be converted from chr to numeric to match other dataframes
str(proftrash_df)
str(gwynndatrash_df)

#converting 
mrtrash_df = 
  mutate(mrtrash_df, year = as.numeric(year))

#binding rows

trashwheels_df = 
  bind_rows(mrtrash_df, proftrash_df, gwynndatrash_df) |>
  janitor::clean_names() 
```

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in June of 2022?

The trashwheel dataset contains data from three trash wheels Mr. Trash Wheel, Professor Trash Wheel and Gwynnda Trash Wheel. There is a total of `r nrow(trashwheels_df)` observations, `r nrow(filter(trashwheels_df,trash_wheel == "Mr. Trash Wheel" ))` observations from Mr. Trash Wheel,`r nrow(filter(trashwheels_df,trash_wheel == "Professor Trash Wheel" ))` observations from Professor Trash Wheel, and `r nrow(filter(trashwheels_df,trash_wheel == "Gwynnda Trash Wheel" ))` observations from Gwynnda Trash Wheel. This dataset includes information on types of trash collected including, `r names(trashwheels_df |> select(plastic_bottles:sports_balls))`, while only sports ball data is only available for Mr. Trash Wheel. Other details on trash collection include `r names(trashwheels_df |> select(date, homes_powered))`, weight in tons and volume in cubic yards.

The total weight of trash collected by Professor Trash Wheel is `r filter(trashwheels_df,trash_wheel == "Professor Trash Wheel") |> select(weight_tons) |>sum() ` tons. 

The total number of cigarette butts collected by Gwynnda in June of 2022 is `r filter(trashwheels_df,trash_wheel == "Gwynnda Trash Wheel", month == "June", year == 2022) |> select(cigarette_butts) |> sum() |> format( scientific=FALSE)`.

#Question 3 


```{r}
mci_baseline_UF = read_csv(file = "./data/MCI_baseline.csv",
 na = c("NA", ".", ""), 
 skip = 1)|> #first line is code book
janitor::clean_names() |>
mutate(sex = recode(sex, `1` = "Male", `0` = "Female"),  #converting to character values
       apoe4= recode(apoe4, `1` = "APOE4 carrier", `0` = "APOE4 non-carrier"))  #converting to character values

mci_baseline = 
  mci_baseline_UF |>
    filter(is.na(age_at_onset) | age_at_onset> current_age)
 # removing individuals who's MCI onset was before study baseline(3 subject), and including NA(those who yet have MCI onset)
```

__Import process__ 

I imported the CSV file, standardized any missing values with "NA", and skipped the first line of the dataset, as it was information on the code book. This allowed the second line to be properly assigned as variable names. I cleaned names and recoded sex and apoe4 from numerical codes (0, 1) to character values, based on the code book from line 1. To ensure only those who met inclusion criteria were included, I then filtered by removing individuals who's MCI onset was before study baseline (`r filter(mci_baseline_UF, age_at_onset<current_age) |> nrow() `) subjects, and included `r filter(mci_baseline, is.na(age_at_onset)) |> nrow()` NA subjects (those who yet have MCI onset). 

__Features of the dataset__

There are `r nrow(mci_baseline)` eligible subjects and `r ncol(mci_baseline)` variables, including: `r names(mci_baseline)`. 

`r nrow(mci_baseline_UF) ` participants were recruited, and `r filter(mci_baseline_UF, !is.na(age_at_onset)) |> nrow()` of these participants develop MCI . 

`r filter(mci_baseline, !is.na(age_at_onset)) |> nrow()` of the `r nrow(mci_baseline)` eligible subjects developed MCI.  

The average baseline age of eligible participants is `r mean(pull(mci_baseline, current_age))|>round()`.

The proportion of women in the study that are APOE4 carriers is `r filter(mci_baseline, sex == "Female" & apoe4 == "APOE4 carrier") |> nrow()` / `r filter(mci_baseline, sex == "Female") |> nrow()` or `r nrow(filter(mci_baseline, sex == "Female" & apoe4 == "APOE4 carrier")) / nrow(filter(mci_baseline, sex == "Female"))` or `r nrow(filter(mci_baseline, sex == "Female" & apoe4 == "APOE4 carrier")) / nrow(filter(mci_baseline, sex == "Female")) * 100` %. 


__Importing dataset of longitudinally observed biomarker values__
```{r}

mci_amyloid = 
read_csv(file = "./data/mci_amyloid.csv",
na = c("NA", ".", ""),
skip = 1) |>  #first line is code book
janitor::clean_names() |> 
rename(id= study_id) 

```



__mci amyloid dataset Import process__ 

I imported the CSV file, standardized any missing values with "NA", and skipped the first line of the dataset, as it was information on the code book. This allowed the second line to be properly assigned as variable names. I cleaned names, and renamed the variable "study ID" to "ID", to match with the ID variable in MCI Baseline dataset.

__mci amyloid dataset Description__

This mci amyloid dataset has `r nrow(mci_amyloid)` subjects and `r ncol(mci_amyloid)` variables, including: `r names(mci_amyloid)`. The timepoint variables measure the time (in years) elapsed since the study baseline to the visit where biomarker Amyloid 42/40 ratio was measured. 


__Joining Datasets__
```{r join}
anti_join(mci_amyloid, mci_baseline, by = "id") |> nrow() # 16 that are in amyloid but not in baseline

anti_join(mci_baseline, mci_amyloid, by = "id") |> nrow() # 8 that are in baseline but not in amyloid

mci_df = 
  inner_join(mci_baseline, mci_amyloid, by = "id") |> view() #selecting inner join for common IDs that appear in both datasets

#exporting final joined cvs 
write_csv(mci_df, "mci_df.csv")

```

__MCI Baseline and MCI Amyloid dataset description__

When comparing the two datasets, there are `r anti_join(mci_amyloid, mci_baseline, by = "id") |> nrow()` participants that are in the amyloid dataset only, but not in the baseline dataset. There are `r anti_join(mci_baseline, mci_amyloid, by = "id") |> nrow()` participants that are in the baseline datasent only, but not in the amyloid dataset. 

When the demographic and biomarker datasets are combined so that only participants who appear in both datasets are retained, there are `r nrow(mci_df)` subjects and `r ncol(mci_df)` combined variables, including: `r names(mci_df)`. 


