---
title: "p8105_hw2_dt2833"
author: "Dominique Tzou"
date: "2026-10-04"
output: github_document


---

instructions: 
Read and clean the data; retain line, station, name, station latitude / longitude, routes served, entry, vending, entrance type, and ADA compliance. Convert the entry variable from character (YES vs NO) to a logical variable (case_match function may be useful).

Write a short paragraph about this dataset – explain briefly what variables the dataset contains, describe your data cleaning steps so far, and give the dimension (rows x columns) of the resulting dataset. Are these data tidy?

Answer the following questions using these data:

How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox); the distinct function may be useful here.
How many stations are ADA compliant?
What proportion of station entrances / exits without vending allow entrance?
Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? Of the stations that serve the A train, how many are ADA compliant?


# Question 1


```{r}
library(tidyverse)
library(dplyr)

transit_df = read_csv(file = "./data/NYC_Transit_Subway_Entrance_And_Exit_Data.csv",
 na = c("NA", ".", ""))|>
janitor::clean_names() |>
select (line, station_name, station_latitude, station_longitude, route1,route2, route3, route4, route5, route6, route7, route8, route9, route10, route11, entry, vending, entrance_type, ada)|> 
  distinct() |>
  mutate(
    entry = recode_values(
      entry,
      "YES"  ~ TRUE,
      "NO" ~ FALSE
    )) #case_match()` was deprecated in dplyr 1.2.0.
```


__Description__

Dataset contains variables: `r names(transit_df)`. For data cleaning, we have cleaned names, selected our desired variables, and standardized missing values to NA. This dataset contains `r ncol(transit_df)` columns and `r nrow(transit_df)` rows. 

How many distinct stations are there? Note that stations are identified both by name and by line (e.g. 125th St 8th Avenue; 125st Broadway; 125st Lenox)? `r n_distinct(pull(transit_df, station_name+line))` unique stations 

This dataset could be more "tidier", the 11 route variable columns are difficult to read 

How many stations are ADA compliant? 84 
What proportion of station entrances / exits without vending allow entrance?
45/117 = 0.3846154

Reformat data so that route number and route name are distinct variables. How many distinct stations serve the A train? 60
Of the stations that serve the A train, how many are ADA compliant? 17

#do i delete NA route name?

```{r ada}
transit_distinct_df = 
  distinct(transit_df, station_name, line, .keep_all = TRUE)  |> nrow()

transit_distinct_df|> 
filter(ada ==TRUE) |> 
  nrow()


n_distinct(pull(transit_df, station_name+line)) # + 
# 472, not 465 
n_distinct(pull(transit_df, interaction(line, station_name)))

transit_df %>%

mutate(station_and_line = interaction(station_name, line)) %>% #interaction: 

pull(station_and_line) %>%

n_distinct()

```
```{r}


```

```{r vending and entry}
filter(transit_df,vending == "NO")|> 
  nrow() #117

filter(transit_df,vending == "NO", entry == "FALSE") |>
  nrow() #72

```

```{r reformat routes}
#how many a train stations
transit_df |>
  mutate (
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)
    ) |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number", 
    names_prefix = "route",
    values_to = "route_name") |>
  
filter(route_name=="A" )|>
 distinct( station_name, line, .keep_all = TRUE) |>
  nrow() #60

#how many are ada compliant
transit_df |>
  mutate (
    route8 = as.character(route8),
    route9 = as.character(route9),
    route10 = as.character(route10),
    route11 = as.character(route11)
    ) |> 
  pivot_longer(
    route1:route11,
    names_to = "route_number", 
    names_prefix = "route",
    values_to = "route_name") |>
  
filter(route_name=="A" )|>
 distinct( station_name, line, .keep_all = TRUE) |>
filter(ada=="TRUE" )|>
nrow()

```

#Question 2 

Importing data from Mr. Trash Wheel, Professor Trash Wheel, Gwynns Falls Trash Wheel.
```{r import data}
library(readxl)

mrtrash_df = read_excel("./data/202509 Trash Wheel Collection Data.xlsx", sheet="Mr. Trash Wheel",  
             na = c("NA", ".", ""))|>
             janitor::clean_names() 
  
proftrash_df = read_excel("./data/202509 Trash Wheel Collection Data.xlsx", sheet="Professor Trash Wheel",
              na = c("NA", ".", ""))|>
             janitor::clean_names() 

gwynnstrash_df = read_excel("./data/202509 Trash Wheel Collection Data.xlsx", sheet="Gwynns Falls Trash Wheel",
              na = c("NA", ".", ""))|>
             janitor::clean_names()  
```

Cleaning Mr. Trash Wheel
```{r}
mrtrash_df = 
  select(mrtrash_df,-x15, -x16)|>  #deleted unnecessary columns 
  filter (!is.na(dumpster)) |> #deleted non-dumpster specific data rows
  mutate(
        sports_balls = round(sports_balls), 
        sports_balls = as.integer(sports_balls), 
        trash_wheel = "Mr. Trash Wheel") |> # creating new variable to indicate wheel source of data 
  relocate(trash_wheel) 

```

Cleaning Professor Trash Wheel
```{r Professor Trash Wheel}
proftrash_df =
  filter (proftrash_df, !is.na(dumpster)) |> #deleted non-dumpster specific data rows
  mutate(trash_wheel = "Professor Trash Wheel")  |> # creating new variable to indicate wheel source of data 
  relocate(trash_wheel) 
```

Cleaning Gwynns Falls Trash Wheel
```{r Gwynns Falls Trash Wheel}
gwynnstrash_df = 
  filter (gwynnstrash_df, !is.na(dumpster)) |> #deleted non-dumpster specific data rows
    mutate(trash_wheel = "Gwynns Falls Trash Wheel") |> # creating new variable to indicate wheel source of data 
  relocate(trash_wheel) 

```

Combining into 1 dataframe
```{r}

str(mrtrash_df) # year variable needs to be converted to numeric to match other dataframes
str(proftrash_df)
str(gwynnstrash_df)

#converting 
mrtrash_df = 
  mutate(mrtrash_df, year = as.numeric(year))

#binding rows

trashwheels_df = 
  bind_rows(mrtrash_df, proftrash_df, gwynnstrash_df) |>
  janitor::clean_names() 
```

Write a paragraph about these data; you are encouraged to use inline R. Be sure to note the number of observations in the resulting dataset, and give examples of key variables. For available data, what was the total weight of trash collected by Professor Trash Wheel? What was the total number of cigarette butts collected by Gwynnda in June of 2022?

The trashwheel dataset contains data from three trash wheels: Mr. Trash Wheel, Professor Trash Wheel and Gwynns Falls Trash Wheel. There is a total of `r nrow(trashwheels_df)` observations, `r nrow(filter(trashwheels_df,trash_wheel == "Mr. Trash Wheel" ))` observations from Mr. Trash Wheel,`r nrow(filter(trashwheels_df,trash_wheel == "Professor Trash Wheel" ))` observations from Professor Trash Wheel, and `r nrow(filter(trashwheels_df,trash_wheel == "Gwynns Falls Trash Wheel" ))` observations from Gwynns Falls Trash Wheel. This dataset includes information on types of trash collected including, `r names(trashwheels_df |> select(plastic_bottles:sports_balls))`, while only sports ball data is only available for Mr. Trash Wheel. Other details on collection include `r names(trashwheels_df |> select(date, homes_powered))`, weight in tones and volume in cubic yards.

The total weight of trash collected by Professor Trash Wheel is `r filter(trashwheels_df,trash_wheel == "Professor Trash Wheel") |> select(weight_tons) |>sum() ` tons. 

The total number of cigarette butts collected by Gwynnda in June of 2022 is `r filter(trashwheels_df,trash_wheel == "Gwynns Falls Trash Wheel", month == "June", year == 2022) |> select(cigarette_butts) |> sum() |> format( scientific=FALSE)`

#Question 3 

Import, clean, and tidy the dataset of baseline demographics. Ensure that sex and APOE4 carrier status are appropriate encoded (i.e. not numeric), and remove any participants who do not meet the stated inclusion criteria (i.e. no MCI at baseline). 

Discuss important steps in the import process and relevant features of the dataset. How many participants were recruited, and of these how many develop MCI? What is the average baseline age? What proportion of women in the study are APOE4 carriers?

Similarly, import, clean, and tidy the dataset of longitudinally observed biomarker values; comment on the steps on the import process and the features of the dataset.

Check whether some participants appear in only the baseline or amyloid datasets, and comment on your findings. Combine the demographic and biomarker datasets so that only participants who appear in both datasets are retained, and briefly describe the resulting dataset; export the result as a CSV to your data directory.

```{r}
mci_baseline_UF = read_csv(file = "./data/MCI_baseline.csv",
 na = c("NA", ".", ""), 
 skip = 1)|> #first line is code book
janitor::clean_names() |>
mutate(sex = recode(sex, `1` = "Male", `0` = "Female"),  #converting to character values
       apoe4= recode(apoe4, `1` = "APOE4 carrier", `0` = "APOE4 non-carrier"))  #converting to character values

mci_baseline = 
  mci_baseline_UF |>
    filter(is.na(age_at_onset) | age_at_onset> current_age)|>
 # removing individuals who's MCI onset was before study baseline(3 subject), including NA(those who yet have MCI onset)
  view()
```

__Import process__ 

I imported the CSV file, standardized any missing values with "NA", and skipped the first line of the dataset, as it was information on the code book. This allowed the second line to be properly assigned as variable names. I cleaned names and recoded sex and apoe4 from numerical codes (0, 1) to character values, based on the code book from line 1. To ensure only those who met inclusion criteria were included, I then filtered by removing individuals who's MCI onset was before study baseline (`r filter(mci_baseline_UF, age_at_onset<current_age) |> nrow() `) subjects), but included `r filter(mci_baseline, is.na(age_at_onset)) |> nrow()` NA subjects (those who yet have MCI onset). 

__features of the dataset__

There are `r nrow(mci_baseline)` subjects and `r ncol(mci_baseline)` variables, including: `r names(mci_baseline)`. 
`r nrow(mci_baseline_UF) ` participants were recruited, `r filter(mci_baseline, !is.na(age_at_onset)) |> nrow()` of these participants develop MCI 

The average baseline age of eligible participants is `r mean(pull(mci_baseline, current_age))`

The proportion of women in the study that are APOE4 carriers is `r filter(mci_baseline, sex == "Female" & apoe4 == "APOE4 carrier") |> nrow()` / `r filter(mci_baseline, sex == "Female") |> nrow()` or `r nrow(filter(mci_baseline, sex == "Female" & apoe4 == "APOE4 carrier")) / nrow(filter(mci_baseline, sex == "Female"))` or `r nrow(filter(mci_baseline, sex == "Female" & apoe4 == "APOE4 carrier")) / nrow(filter(mci_baseline, sex == "Female")) * 100` %


```{r}
mci_amyloid = read_csv(file = "./data/mci_amyloid.csv",
 na = c("NA", ".", ""),
 skip = 1)|>  #first line is code book
janitor::clean_names() |> 
rename(id= study_id) |> view()
```



__mci amyloid dataset Import process__ 

I imported the CSV file, standardized any missing values with "NA", and skipped the first line of the dataset, as it was information on the code book. This allowed the second line to be properly assigned as variable names. I cleaned names, and renamed the variable study ID to ID, to match with the ID variable in MCI Baseline

__mci amyloid dataset Description__

This mci amyloid dataset has `r nrow(mci_amyloid)` subjects and `r ncol(mci_amyloid)` variables, including: `r names(mci_amyloid)`. The timepoint variables measure the time (in years) elapsed since the study baseline to the visit where biomarker Amyloid _ 42/40 ratio was measured. 

```{r}
anti_join(mci_amyloid, mci_baseline, by = "id") |> nrow() # 16 that are in amyloid but not in baseline

anti_join(mci_baseline, mci_amyloid, by = "id") |> nrow() # 8 that are in baseline but not in amyloid

mci_df = 
  inner_join(mci_baseline, mci_amyloid, by = "id") |> view()

#exporting final joined cvs 
write_csv(mci_df, "mci_df.csv")

```

__MCI Baseline and MCI Amyloid dataset description__

When comparing the two datasets, there are `r anti_join(mci_amyloid, mci_baseline, by = "id") |> nrow()` participants that are in the amyloid dataset only, but not in the baseline dataset. There are `r anti_join(mci_baseline, mci_amyloid, by = "id") |> nrow()` participants that are in the baseline datasent only, but not in the amyloid dataset. 

When the demographic and biomarker datasets are combined so that only participants who appear in both datasets are retained, there are `r nrow(mci_df)` subjects and `r ncol(mci_df)` combined variables, including: `r names(mci_df)`. 


